### 2017-05-17 paper/content/topologySkeletons.tex
renamed \code to \inlinecode

\FloatBarrier
\section{Topological Skeletons}
\label{sec:topology-skeletons}
Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons. The Eden library leverages this problem and already comes with more predefined skeletons, among them a \code{pipe}, a \code{ring} and a \code{torus} implementation \cite{Loogen2012, eden_skel_topology}. These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.

@@

The parallel pipe skeleton is semantically equivalent to folding over a list \code{[arr a a]} of arrows with \code{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the \code{ArrowLoop} typeclass which gives us the \code{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\mbcomment{das kann man hier so lassen, oder?}
\begin{lstlisting}[frame=htrbl]
loop (arr (\(a, b) -> (b, a:b)))
\end{lstlisting}
, which is the same as
\begin{lstlisting}[frame=htrbl]
loop (arr snd &&& arr (uncurry (:)))
\end{lstlisting}
defines an arrow that takes its input \code{a} and converts it into an infinite stream \code{[a]} of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \code{parEvalN conf fs} inside the second argument of \code{(\&\&\&)} and then only picking the first element of the resulting list with \code{arr last}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
	arr last

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{lstlisting}
\caption{A first draft of the pipe skeleton expressed with parallel arrows. Note that the use of \code{lazy} is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input \code{[a]} terminates before passing it into \code{parEvalN}.}
\label{fig:pipeSimple}
\end{figure}

However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}. Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \code{pipe} \ref{fig:pipe}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
	Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{lstlisting}
\caption{Final definition of the pipe skeleton which uses Futures}
\label{fig:pipe}
\end{figure}

Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. \code{arr a b} and \code{arr b c}. By wrapping these two arrows inside a common type we obtain \code{pipe2} (Fig.~\ref{fig:pipe2}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipe2 :: (ArrowLoop arr, ArrowChoice arr,
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
	Future fut (([a], [b]), [c])) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g = 
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>>
	arr head
		where
			unify :: (ArrowChoice arr) =>
				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))
\end{lstlisting}
\caption{Definition of pipe2}
\label{fig:pipe2}
\end{figure}

Note that extensive use of this combinator over \code{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN. Nonetheless, we can define a parallel piping operator \code{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \code{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|>>>|) :: (ArrowLoop arr, ArrowChoice arr,
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
	Future fut (([a], [b]), [c])) =>
	arr a b -> arr b c -> arr a c
(|>>>|) = pipe2 ()
\end{lstlisting}
\caption{Definition of \texttt{(|>>>|)}}
\label{fig:|>>>|}
\end{figure}

\subsection{Ring skeleton} \label{sec:ring}
\begin{figure}[h]
	\includegraphics[scale=0.75]{images/ring}
	\caption{Schematic depiction of the ring skeleton}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \code{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other. Its input is a node function \code{i -> r -> (o, r)} in which \code{r} serves as the intermediary output that gets send to the neighbour of each node. This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
ringSimple :: (Trans i, Trans o, Trans r) =>
   (i -> r -> (o,r))
   -> [i] -> [o]
ringSimple f is =  os
  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
                                   (zip is $ lazy ringIns))
        ringIns = rightRotate ringOuts

@@

We can rewrite its functionality easily with the use of \code{loop} as the definition of the node function, \code{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \code{loop}'s \code{arr (a, b) (c, b) -> arr a c}. In each iteration we start by rotating the intermediary input from the nodes \code{[fut r]} with \code{second (rightRotate >>> lazy)}. Similarly to the \code{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \code{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
%\begin{quotation}
\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate. Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
%\end{quotation}
Next, we zip the resulting \code{([i], [fut r])} to \code{[(i, fut r)]} with \code{arr (uncurry zip)} so we can feed that into a our input arrow \code{arr (i, r) (o, r)}, which we transform into \code{arr (i, fut r) (o, fut r)} before lifting it to \code{arr [(i, fut r)] [(o, fut r)]} to get a list \code{[(o, fut r)]}. Finally we unzip this list into \code{([o], [fut r])}. Plugging this arrow \code{arr ([i], [fut r]) ([o], fut r)} into the definition of \code{loop} from earlier gives us \code{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
ring :: (ArrowLoop arr, Future fut r,
	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
    conf ->
    arr (i, r) (o, r) ->
    arr [i] [o]
ring conf f =
	loop (second (rightRotate >>> lazy) >>>
    arr (uncurry zip) >>>
    parMap conf (second get >>> f >>> second put) >>>
    arr unzip)

@@

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{lstlisting}
\caption{Final Definition of the ring Skeleton}
\label{fig:ringFinal}
\end{figure}
 %$ %% formatting
% and lazy:
% \begin{lstlisting}[frame=htrbl]			
% lazy :: (Arrow arr) => arr [a] [a]
% lazy = arr (\ ~(x:xs) -> x : lazy xs
% \end{lstlisting}
%% we have shown this already.
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. 
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}
\subsection{Torus skeleton}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the torus skeleton}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
\\\\
With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \code{ArrowLoop} typeclass.
\\\\
Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple \code{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \code{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \code{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \code{[[fut a]]} horizontally and \code{[[fut b]]} vertically. Then, we once again zip the inputs for the input arrow with \code{arr (uncurry3 zipWith3 lazyzip3)} from \code{([[c]], ([[fut a]], [[fut b]]))} to \code{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
\\\\
This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to \code{shuffle} all the inputs to then pass it into \code{parMap conf (ptorus f)} which yields us \code{[(d, fut a, fut b)]}. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \code{arr length >>> arr unshuffle} with the use of \code{app :: arr (arr a b, a) c} from the \code{ArrowApply} typeclass. Finally, we unpack this matrix \code{[[[(d, fut a, fut b)]]} with \code{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \code{([[d]], ([[fut a]], [[fut b]]))}.
\\\\
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
	Future fut a, Future fut b) =>
	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
torus conf f =
	loop (second ((mapArr rightRotate >>> lazy) ***
		(arr rightRotate >>> lazy)) >>>
	arr (uncurry3 (zipWith3 lazyzip3)) >>>
	(arr length >>> arr unshuffle) &&&
		(shuffle >>> parMap conf (ptorus f)) >>>
	app >>>
	arr (map unzip3) >>> arr unzip3 >>> threetotwo)

>>>>

\FloatBarrier
\section{Topological Skeletons}
\label{sec:topology-skeletons}
Even though many algorithms can be expressed by parallel maps, some problems require more sophisticated skeletons. The Eden library leverages this problem and already comes with more predefined skeletons, among them a \inlinecode{pipe}, a \inlinecode{ring} and a \inlinecode{torus} implementation \cite{Loogen2012, eden_skel_topology}. These seem like reasonable candidates to be ported to our arrow based parallel Haskell to showcase that we can express more sophisticated skeletons with Parallel Arrows as well.

@@

The parallel pipe skeleton is semantically equivalent to folding over a list \inlinecode{[arr a a]} of arrows with \inlinecode{>>>}, but does this in parallel, meaning that the arrows do not have to reside on the same thread/machine. We implement this skeleton using the \inlinecode{ArrowLoop} typeclass which gives us the \inlinecode{loop :: arr (a, b) (c, b) -> arr a c} combinator which allows us to express recursive fix-point computations in which output values are fed back as input. For example %this
\mbcomment{das kann man hier so lassen, oder?}
\begin{lstlisting}[frame=htrbl]
loop (arr (\(a, b) -> (b, a:b)))
\end{lstlisting}
, which is the same as
\begin{lstlisting}[frame=htrbl]
loop (arr snd &&& arr (uncurry (:)))
\end{lstlisting}
defines an arrow that takes its input \inlinecode{a} and converts it into an infinite stream \inlinecode{[a]} of it. Using this to our advantage gives us a first draft of a pipe implementation (Fig.~\ref{fig:pipeSimple}) by plugging in the parallel evaluation call \inlinecode{parEvalN conf fs} inside the second argument of \inlinecode{(\&\&\&)} and then only picking the first element of the resulting list with \inlinecode{arr last}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipeSimple :: (ArrowLoop arr, ArrowParallel arr a a conf) =>
	conf -> [arr a a] -> arr a a
pipeSimple conf fs =
	loop (arr snd &&&
		(arr (uncurry (:) >>> lazy) >>> parEvalN conf fs)) >>>
	arr last

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{lstlisting}
\caption{A first draft of the pipe skeleton expressed with parallel arrows. Note that the use of \inlinecode{lazy} is essential as without it programs using this definition would never halt. We need to enforce that the evaluation of the input \inlinecode{[a]} terminates before passing it into \inlinecode{parEvalN}.}
\label{fig:pipeSimple}
\end{figure}

However, using this definition directly will make the master node a potential bottleneck in distributed environments as described in Section~\ref{futures}. Therefore, we introduce a more sophisticated version that internally uses Futures and get the final definition of \inlinecode{pipe} \ref{fig:pipe}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipe :: (ArrowLoop arr, ArrowParallel arr (fut a) (fut a) conf,
	Future fut a) =>
	conf -> [arr a a] -> arr a a
pipe conf fs = unliftFut (pipeSimple conf (map liftFut fs))
\end{lstlisting}
\caption{Final definition of the pipe skeleton which uses Futures}
\label{fig:pipe}
\end{figure}

Sometimes, this pipe definition can be a bit inconvenient, especially if we want to pipe arrows of mixed types together, i.e. \inlinecode{arr a b} and \inlinecode{arr b c}. By wrapping these two arrows inside a common type we obtain \inlinecode{pipe2} (Fig.~\ref{fig:pipe2}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
pipe2 :: (ArrowLoop arr, ArrowChoice arr,
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) conf,
	Future fut (([a], [b]), [c])) =>
	conf -> arr a b -> arr b c -> arr a c
pipe2 conf f g = 
	(arr return &&& arr (const [])) &&& arr (const []) >>>
	pipe conf (replicate 2 (unify f g)) >>>
	arr snd >>>
	arr head
		where
			unify :: (ArrowChoice arr) =>
				arr a b -> arr b c -> arr (([a], [b]), [c]) (([a], [b]), [c])
			unify f g =
				(mapArr f *** mapArr g) *** arr (\_ -> []) >>>
				arr (\((a, b), c) -> ((c, a), b))
\end{lstlisting}
\caption{Definition of pipe2}
\label{fig:pipe2}
\end{figure}

Note that extensive use of this combinator over \inlinecode{pipe} with a hand-written combination data type will probably result in worse performance because of more communication overhead from the many calls to parEvalN. Nonetheless, we can define a parallel piping operator \inlinecode{(|>>>|)} (Fig.~\ref{fig:|>>>|}, which is semantically equivalent to \inlinecode{(>>>)} in a similar manner to the other parallel syntactic sugar from section~\ref{syntacticSugar}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|>>>|) :: (ArrowLoop arr, ArrowChoice arr,
	ArrowParallel arr (fut (([a], [b]), [c])) (fut (([a], [b]), [c])) (),
	Future fut (([a], [b]), [c])) =>
	arr a b -> arr b c -> arr a c
(|>>>|) = pipe2 ()
\end{lstlisting}
\caption{Definition of \texttt{(|>>>|)}}
\label{fig:|>>>|}
\end{figure}

\subsection{Ring skeleton} \label{sec:ring}
\begin{figure}[h]
	\includegraphics[scale=0.75]{images/ring}
	\caption{Schematic depiction of the ring skeleton}
	\label{fig:ringImg}
\end{figure}
Eden comes with a ring skeleton (Fig.~\ref{fig:ringImg}) implementation that allows the computation of a function \inlinecode{[i] -> [o]} with a ring of nodes that communicate in a ring topology with each other. Its input is a node function \inlinecode{i -> r -> (o, r)} in which \inlinecode{r} serves as the intermediary output that gets send to the neighbour of each node. This data is sent over direct communication channels (remote data) (Fig.~\ref{fig:ringEden}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
ringSimple :: (Trans i, Trans o, Trans r) =>
   (i -> r -> (o,r))
   -> [i] -> [o]
ringSimple f is =  os
  where (os,ringOuts) = unzip (parMap (toRD $ uncurry f)
                                   (zip is $ lazy ringIns))
        ringIns = rightRotate ringOuts

@@

We can rewrite its functionality easily with the use of \inlinecode{loop} as the definition of the node function, \inlinecode{arr (i, r) (o, r)}, after being transformed into an arrow, already fits quite neatly into the \inlinecode{loop}'s \inlinecode{arr (a, b) (c, b) -> arr a c}. In each iteration we start by rotating the intermediary input from the nodes \inlinecode{[fut r]} with \inlinecode{second (rightRotate >>> lazy)}. Similarly to the \inlinecode{pipe} (Fig.~\ref{fig:pipeSimple},~\ref{fig:pipe}), we have to feed the intermediary input into our \inlinecode{lazy} arrow here, or the evaluation would hang.\olcomment{meh, wording} The reasoning is explained by \citet{Loogen2012}:
%\begin{quotation}
\enquote{Note that the list of ring inputs ringIns is the same as the list of ring outputs ringOuts rotated by one element to the right using the auxiliary function rightRotate. Thus, the program would get stuck without the lazy pattern, because the ring input will only be produced after process creation and process creation will not occur without the first input.}
%\end{quotation}
Next, we zip the resulting \inlinecode{([i], [fut r])} to \inlinecode{[(i, fut r)]} with \inlinecode{arr (uncurry zip)} so we can feed that into a our input arrow \inlinecode{arr (i, r) (o, r)}, which we transform into \inlinecode{arr (i, fut r) (o, fut r)} before lifting it to \inlinecode{arr [(i, fut r)] [(o, fut r)]} to get a list \inlinecode{[(o, fut r)]}. Finally we unzip this list into \inlinecode{([o], [fut r])}. Plugging this arrow \inlinecode{arr ([i], [fut r]) ([o], fut r)} into the definition of \inlinecode{loop} from earlier gives us \inlinecode{arr [i] [o]}, our ring arrow (Fig.~\ref{fig:ringFinal}).
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
ring :: (ArrowLoop arr, Future fut r,
	ArrowParallel arr (i, fut r) (o, fut r) conf) =>
    conf ->
    arr (i, r) (o, r) ->
    arr [i] [o]
ring conf f =
	loop (second (rightRotate >>> lazy) >>>
    arr (uncurry zip) >>>
    parMap conf (second get >>> f >>> second put) >>>
    arr unzip)

@@

lazy :: (Arrow arr) => arr [a] [a]
lazy = arr (\ ~(x:xs) -> x : lazy xs)
\end{lstlisting}
\caption{Final Definition of the ring Skeleton}
\label{fig:ringFinal}
\end{figure}
 %$ %% formatting
% and lazy:
% \begin{lstlisting}[frame=htrbl]			
% lazy :: (Arrow arr) => arr [a] [a]
% lazy = arr (\ ~(x:xs) -> x : lazy xs
% \end{lstlisting}
%% we have shown this already.
This combinator can, for example, be used to calculate the shortest paths in a graph using Warshall's algorithm. 
%Further details on this can be found in \cite{eden_cefp}.
\olcomment{let's do it?}
\subsection{Torus skeleton}
\begin{figure}
	\includegraphics[scale=0.75]{images/torus}
	\caption{Schematic depiction of the torus skeleton}
	\label{fig:ringTorusImg}
\end{figure}
If we take the concept of a ring from \ref{sec:ring} one dimension further, we get a torus (Fig.~\ref{fig:ringTorusImg},~\ref{fig:torus}). Every node sends ands receives data from horizontal and vertical neighbours in each communication round.
\\\\
With our parallel Arrows we re-implement the torus combinator from Eden \citep{eden_skel_topology} - yet again with the help of the \inlinecode{ArrowLoop} typeclass.
\\\\
Similar to the ring, we once again start by rotating the input, but this time not only in one direction, but in two. This means that the intermediary input from the neighbour nodes has to be stored in a tuple \inlinecode{([[fut a]], [[fut b]])} in the second argument (loop only allows for 2 arguments) of our looped arrow \inlinecode{arr ([[c]], ([[fut a]], [[fut b]])) ([[d]], ([[fut a]], [[fut b]]))} and our rotation arrow becomes \inlinecode{second ((mapArr rightRotate >>> lazy) *** (arr rightRotate >>> lazy))} instead of the singular rotation in the ring as we rotate \inlinecode{[[fut a]]} horizontally and \inlinecode{[[fut b]]} vertically. Then, we once again zip the inputs for the input arrow with \inlinecode{arr (uncurry3 zipWith3 lazyzip3)} from \inlinecode{([[c]], ([[fut a]], [[fut b]]))} to \inlinecode{[[(c, fut a, fut b)]]}, which we then feed into our parallel execution.
\\\\
This, however, is more complicated than in the ring case as we have one more dimension of inputs to be transformed. We first have to \inlinecode{shuffle} all the inputs to then pass it into \inlinecode{parMap conf (ptorus f)} which yields us \inlinecode{[(d, fut a, fut b)]}. We can then unpack this shuffled list back to its original ordering by feeding it into the specific unshuffle arrow we created one step earlier with \inlinecode{arr length >>> arr unshuffle} with the use of \inlinecode{app :: arr (arr a b, a) c} from the \inlinecode{ArrowApply} typeclass. Finally, we unpack this matrix \inlinecode{[[[(d, fut a, fut b)]]} with \inlinecode{arr (map unzip3) >>> arr unzip3 >>> threetotwo} to get  \inlinecode{([[d]], ([[fut a]], [[fut b]]))}.
\\\\
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
torus :: (ArrowLoop arr, ArrowChoice arr, ArrowApply arr,
	ArrowParallel arr (c, fut a, fut b) (d, fut a, fut b) conf,
	Future fut a, Future fut b) =>
	conf -> arr (c, a, b) (d, a, b) -> arr [[c]] [[d]]
torus conf f =
	loop (second ((mapArr rightRotate >>> lazy) ***
		(arr rightRotate >>> lazy)) >>>
	arr (uncurry3 (zipWith3 lazyzip3)) >>>
	(arr length >>> arr unshuffle) &&&
		(shuffle >>> parMap conf (ptorus f)) >>>
	app >>>
	arr (map unzip3) >>> arr unzip3 >>> threetotwo)

### 2017-05-17 paper/content/utilityFunctions.tex
renamed \code to \inlinecode

\section{Utility Functions}\label{utilfns}
To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \code{map}, \code{foldl} and \code{zipWith} on arrows.

The \code{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \code{arr a b} to an arrow \code{arr [a] [b]} \cite{programming_with_arrows}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
mapArr f =
	arr listcase >>>
	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
	where listcase [] = Left ()
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\code{map} over arrows}
\label{fig:mapArr}
\end{figure}
Similarly, we can also define \code{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \code{arr (b, a) b} with a neutral element \code{b} to \code{arr [a] b}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
foldlArr f b =
	arr listcase >>>
	arr (const b) |||
		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
	where listcase [] = Left []
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\code{foldl} over arrows}
\label{fig:foldlArr}
\end{figure}
Finally, with the help of \code{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \code{arr (a, b) c} to an arrow \code{arr ([a], [b]) [c]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
\end{lstlisting}
\caption{\code{zipWith} over arrows}
\label{fig:zipWithArr}
\end{figure}
 %$ %% formatting
These combinators make use of the \code{ArrowChoice} type class which provides the \code{(|||)} combinator. It takes two arrows \code{arr a c} and \code{arr b c} and combines them into a new arrow \code{arr (Either a b) c} which pipes all \code{Left a}'s to the first arrow and all \code{Right b}'s to the second arrow.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
\end{lstlisting}
\caption{Type signature of \code{(|||)}}
\label{fig:codeSig|||}
\end{figure}

With the zipWithArr combinator we can also write a combinator \code{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \code{[arr a b]} to an arrow \code{arr [a] [b]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
\end{lstlisting}
\caption{Definition of \code{listApp}}
\label{fig:listApp}
\end{figure}
% $ %% formatting
. Note that  this additionally makes use of the \code{ArrowApply} typeclass that allows us to evaluate arrows with \code{app :: arr (arr a b, a) c}.

>>>>

\section{Utility Functions}\label{utilfns}
To be able to go into detail on parallel arrows, we introduce some utility combinators first, that will help us later: \inlinecode{map}, \inlinecode{foldl} and \inlinecode{zipWith} on arrows.

The \inlinecode{mapArr} combinator (Fig.~\ref{fig:mapArr}) lifts any arrow \inlinecode{arr a b} to an arrow \inlinecode{arr [a] [b]} \cite{programming_with_arrows}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
mapArr :: ArrowChoice arr => arr a b -> arr [a] [b]
mapArr f =
	arr listcase >>>
	arr (const []) ||| (f *** mapArr f >>> arr (uncurry (:)))
	where listcase [] = Left ()
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\inlinecode{map} over arrows}
\label{fig:mapArr}
\end{figure}
Similarly, we can also define \inlinecode{foldlArr} (Fig.~\ref{fig:foldlArr}) that lifts any arrow \inlinecode{arr (b, a) b} with a neutral element \inlinecode{b} to \inlinecode{arr [a] b}:
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
foldlArr :: (ArrowChoice arr, ArrowApply arr) => arr (b, a) b -> b -> arr [a] b
foldlArr f b =
	arr listcase >>>
	arr (const b) |||
		(first (arr (\a -> (b, a)) >>> f >>> arr (foldlArr f)) >>> app)
	where listcase [] = Left []
	      listcase (x:xs) = Right (x,xs)
\end{lstlisting}
\caption{\inlinecode{foldl} over arrows}
\label{fig:foldlArr}
\end{figure}
Finally, with the help of \inlinecode{mapArr} (Fig.~\ref{fig:mapArr}), we can define \lstinline{zipWithArr} (Fig.~\ref{fig:zipWithArr}) that lifts any arrow \inlinecode{arr (a, b) c} to an arrow \inlinecode{arr ([a], [b]) [c]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
zipWithArr :: ArrowChoice arr => arr (a, b) c -> arr ([a], [b]) [c]
zipWithArr f = (arr $ \(as, bs) -> zipWith (,) as bs) >>> mapArr f
\end{lstlisting}
\caption{\inlinecode{zipWith} over arrows}
\label{fig:zipWithArr}
\end{figure}
 %$ %% formatting
These combinators make use of the \inlinecode{ArrowChoice} type class which provides the \inlinecode{(|||)} combinator. It takes two arrows \inlinecode{arr a c} and \inlinecode{arr b c} and combines them into a new arrow \inlinecode{arr (Either a b) c} which pipes all \inlinecode{Left a}'s to the first arrow and all \inlinecode{Right b}'s to the second arrow.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
(|||) :: ArrowChoice arr a c -> arr b c -> arr (Either a b) c
\end{lstlisting}
\caption{Type signature of \inlinecode{(|||)}}
\label{fig:codeSig|||}
\end{figure}

With the zipWithArr combinator we can also write a combinator \inlinecode{listApp} (Fig.~\ref{fig:listApp}), that lifts a list of arrows \inlinecode{[arr a b]} to an arrow \inlinecode{arr [a] [b]}.
\begin{figure}[h]
\begin{lstlisting}[frame=htrbl]
listApp :: (ArrowChoice arr, ArrowApply arr) => [arr a b] -> arr [a] [b]
listApp fs = (arr $ \as -> (fs, as)) >>> zipWithArr app
\end{lstlisting}
\caption{Definition of \inlinecode{listApp}}
\label{fig:listApp}
\end{figure}
% $ %% formatting
. Note that  this additionally makes use of the \inlinecode{ArrowApply} typeclass that allows us to evaluate arrows with \inlinecode{app :: arr (arr a b, a) c}.
